---
title: "R Data Frames"
author: "Robert Barr"
date: "1/11/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# The R Data Frame

## Tutorial Overview:
This tutorial will guide and teach you some R data frame fundamentals. First you will create vectors which will act as your base data source for creating your vector. Then you will learn how to add rows and columns to an R data frame. You will learn how to create a calculated field using implied looping. You will then learn how to query and sort an R data frame. Finally you will learn how to update values and drop rows.


## General thoughts & notes
- A 2 dimensional spreadsheet-like data structure similar to a database table
   - Think rows X columns
- An important R construct used R data structure in data science and statistical analysis
- Can be thought of as a special kind of list:
  - Can store various data types such as numbers, text, and logical variables
  - **CRITICAL**: Data frame members must all be equal length vectors.
  
## Links:
- [data.frame()](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/data.frame) to create a data frame
  - [R CODER](https://r-coder.com/dataframe-r/) article, "DataFrame in R"
  - [stringAsFactors = TRUE(default) | FALSE](https://developer.r-project.org/Blog/public/2020/02/16/stringsasfactors/index.html)
- [str()](https://www.rdocumentation.org/packages/utils/versions/3.6.2/topics/str) function for data frame structure
- [dim()](https://r-lang.com/dim-in-r/) function for data frame dimensions rowsXcols
- [colnames()](https://rdrr.io/r/base/colnames.html) function to list column names
- [head()]() function ouput the first n number of rows
- [tail()]() function output the last n number of rows
- [summary()]() function output summary statistics
- [cbind()]() function used to add columns
- [rbind()]() function used to add rows
- [order()](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/order) function used to sort and order data
- [Datacamp.com's Tutorial on the R Apply Family](https://www.datacamp.com/community/tutorials/r-tutorial-apply-family)

# Example:
## Create the base vector data sources
These vectors will serve as the initial data sources of our data frame.

```{r base vectors, include=T, error=F, warning=F}
v_student_ids <- c(1001, 1002, 1003)
v_first_names <- c("John", "Mary", "Tyra")
v_last_names <- c("Smith", "Jones", "Washington")
v_email_addresses <- c("john.smith@test.com", "mj@example.com", "washtyra@test.info")
```

## Create the class roster data frame
Use the previously created vectors to create a data frame.

```{r create data frame, include=T, error=F, warning=F}
df_MAT101_Roster <- data.frame(StudentID = v_student_ids, FirstName = v_first_names, LastName = v_last_names, Email = v_email_addresses, stringsAsFactors = FALSE)
str(df_MAT101_Roster)
colnames(df_MAT101_Roster)
sapply(df_MAT101_Roster, class)
dim(df_MAT101_Roster)
```


## Add Additional Rows & Cols
You can add additional rows and columns to an existing data frame.

```{r add rows & cols, include = T, error = F, waring = F}
df_MAT101_Roster <- rbind(df_MAT101_Roster, c(1004, "Fred", "Johnson", "Fred.Johnson@test.us"))

df_MAT101_Roster <- cbind(df_MAT101_Roster, Test1 = c(87, 93, 92, 76) )
df_MAT101_Roster <- cbind(df_MAT101_Roster, Test2 = c(89, 95, 91, 78) )
df_MAT101_Roster <- cbind(df_MAT101_Roster, Test3 = c(86, 91, 92, 75) )
df_MAT101_Roster <- cbind(df_MAT101_Roster, Test4 = c(88, 90, 93, 76) )
df_MAT101_Roster <- cbind(df_MAT101_Roster, Final = c(88, 92, 94, 74) )

```

## Creating calculated fields: Total Points & Avg
You can derive new columns based on calculations. The [apply family of functions](https://www.datacamp.com/community/tutorials/r-tutorial-apply-family) uses implict row iteration to apply functions to row values. 

```{r create calculated fields, include=TRUE, error=FALSE, warning=FALSE}
df_MAT101_Roster$TotalScore <- apply(X = df_MAT101_Roster[, c(5:9)], MARGIN = 1, FUN = sum)
df_MAT101_Roster$Avg <- round(apply( X = df_MAT101_Roster[, c(5:9)], MARGIN = 1, FUN = mean ), 2)
```

## Create the score to letter grade function
```{r custom function, include=TRUE, error=FALSE, warning=FALSE}
f_to_letter_grade <- function(score){
  if(score >= 90 && score <= 100){
    return('A')
  } else if( score >= 80 && score < 90){
    return('B')
  } else if (score >= 70 && score < 80){
    return('C')
  } else if (score >= 60 && score < 70){
    return('D')
  } else if(score < 60){
    return('F')
  } else {
    return('N/A')
  }
  
}
```


## apply the letter grade function & convert to factor
```{r letter grade factor, include=TRUE, warning=FALSE, error=FALSE}
df_MAT101_Roster <- cbind( df_MAT101_Roster, 'Final Letter Grade' = sapply( X = df_MAT101_Roster$Avg, f_to_letter_grade) )
df_MAT101_Roster$`Final Letter Grade` <- factor(df_MAT101_Roster$`Final Letter Grade`)
```

## Add additional students (rows)
```{r add more students, include=TRUE, warning=FALSE, error=FALSE}
df_MAT101_Roster <- rbind(df_MAT101_Roster, list(1010, "Jim", "Smith", "jsmith@test.edu", 88, 90, 86, 85, 80, 0, 0, NA))
df_MAT101_Roster <- rbind(df_MAT101_Roster, list(1011, "George", "Lockwood", "lockwood@test.edu", 77, 79, 86, 75, 68, 0, 0, NA))
df_MAT101_Roster <- rbind(df_MAT101_Roster, list(1012, "Frank", "Xavier", "frank-o@test.edu", 92, 90, 86, 85, 81, 0, 0, NA))
```

### Calculate Total Score, Avg, and Letter Grade for new students
```{r re-calculate for new students, include=TRUE, warning=FALSE, error=FALSE}
df_MAT101_Roster$TotalScore <- apply(X = df_MAT101_Roster[, c(5:9)], MARGIN = 1, FUN = sum)
df_MAT101_Roster$Avg <- round(apply( X = df_MAT101_Roster[, c(5:9)], MARGIN = 1, FUN = mean ), 2)
df_MAT101_Roster[ df_MAT101_Roster["StudentID"] >= 1010, ]$`Final Letter Grade` = sapply(X = df_MAT101_Roster[ df_MAT101_Roster["StudentID"] >= 1010, ]$Avg, f_to_letter_grade)

dim(df_MAT101_Roster)
```

## Reporting

### What's the letter grade distribution?
```{r queries and reports, include = TRUE, error = FALSE, warning = FALSE}
table(df_MAT101_Roster$`Final Letter Grade`)
```

### Who has an A in the class
```{r letter grade report, include=TRUE, error = FALSE, warning = FALSE}

df_MAT101_Roster[df_MAT101_Roster$`Final Letter Grade` == "A", ]

```

### Sort by GPA
```{r sorting, include=TRUE, error=FALSE, warning=FALSE}
df_MAT101_Roster <- df_MAT101_Roster[ order(df_MAT101_Roster$Avg, decreasing = TRUE), ]

df_MAT101_Roster
```

## Update Values
### Change a "C" to "W"
```{r update letter grade, include=T, error=F, warning=F}
df_MAT101_Roster$`Final Letter Grade` <- as.character(df_MAT101_Roster$`Final Letter Grade`)
df_MAT101_Roster$`Final Letter Grade`[df_MAT101_Roster$`Final Letter Grade` == 'C'] <- 'W'
df_MAT101_Roster$`Final Letter Grade` <- factor(df_MAT101_Roster$`Final Letter Grade`)

table(df_MAT101_Roster$`Final Letter Grade`)
```

## Drop Rows
```{r drop rows, include=T, error = F, warning=F}
df_MAT101_Roster_Final <- df_MAT101_Roster[!df_MAT101_Roster$`Final Letter Grade` == "W" ,]
table(df_MAT101_Roster_Final$`Final Letter Grade`)

df_MAT101_Roster_Final <- df_MAT101_Roster_Final[ order(df_MAT101_Roster_Final$LastName, decreasing = FALSE), ]
df_MAT101_Roster_Final
```





